Stressing Bank Profitability for Interest Rate Risk$^*$

### 2013-07-09 LATEX/beg.draft.2013.tex
updated draft

\begin{document}
\renewcommand{\baselinestretch}{1}
\author{{\Large Valentin Bolotnyy} \\
Federal Reserve Board\thanks{E-mail: valentin.bolotnyy@frb.gov.\vspace{0.05in}} \and
{\Large Rochelle M. Edge} \\
Federal Reserve Board\thanks{E-mail: rochelle.m.edge@frb.gov.\vspace{0.05in}}\and
{\Large Luca Guerrieri} \\
\addtocounter{footnote}{5} Federal Reserve Board\thanks{%
E-mail: luca.guerrieri@frb.gov.}}
\title{\textbf{Stressing Bank Profitability for Interest Rate Risk}}
\date{\today}
\maketitle

\begin{abstract}
\vspace{-0.06in} \noindent  Regular bank stress testing (and the capital planning that it implies) is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime.
 The form of stress testing mandated under the Dodd-Frank Act for the Federal Reserve to undertake and the form being used by the Federal Reserve for its annual Comprehensive Capital Analysis and Review (CCAR) is one based around a small number of scenarios.  This approach involves specifying macroeconomic and financial scenarios that represents stressful conditions over the time horizon of the stress test and then projecting, for each BHC in the stress test, its losses, income, and path of capital \textit{given that scenario}. In this paper, we focus on the revenue generate aspect of banks and consider the ability of empirical models to translate developments in interest rates into bank net interest margins (NIMs). We consider an extensive number of time-series models. While our best-performing models beat a no-change forecast, the confidence bands are so large that baseline and stress scenarios taken from CCAR appear indistinguishable.

@@

\noindent Key Words:

@@

\noindent JEL Classification: To be completed.

@@

\vspace{-0.1in}
Regular bank stress testing (and the capital planning that it implies) is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime.\renewcommand{\baselinestretch}{1}\footnote{See Governor Tarullo’s November 2011 speech ``The Evolution of Capital Regulation'' for this characterization of stress testing. The two other reforms that he mentions are improvements to the traditional, individual firm approach to capital regulation and the introduction of a macroprudential component to capital regulation.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} While stress testing can take many forms---for example, being based on a small number of scenarios, being based on a large number of scenarios, or even being undertaken in reverse form to achieve a certain adverse outcome---the form of the stress tests mandated under the Dodd-Frank Act for the Federal Reserve to undertake and the form being used by the Federal Reserve for its annual Comprehensive Capital Analysis and Review (CCAR) is one based around a small number of scenarios.  This approach involves specifying a small number of macroeconomic and financial scenarios that represents stressful conditions over the time horizon of the stress test and then projecting, for each BHC in the stress test, its losses, income, and path of capital \textit{given that scenario}. Clearly, there are at least two important parts to conducting bank stress tests (i)~formulating macroeconomic and financial scenarios that should be stressful to banks and (ii)~developing econometric models that are able to translate such stressful conditions into losses, income, and capital.
%This paper considers the latter of these two issues; that is, translating
%stressful economic conditions into bank income.

In developing econometric models to translate stressful economic conditions into key bank financial-statement variables it is important to consider what types of developments are likely to be stressful to banks and what particular economic variables are likely to capture these stresses. It is then critical that the models being used for stress testing are able to translate developments in these economic variables into the key bank financial-statement variables that the model is designed to project.  In this paper, we examine this property for models of bank net interest margins (NIMs), defined as the ratio of net interest income (NII) to interest earning assets, where NII is the difference between a bank's interest income and interest expenses.

Our motivation for focusing on models of NIMs is based on a number of considerations. First, bank losses that arise from banks' having to pay out more in interest on their liabilities than they are earning from their assets should not be overlooked as an important source of risk to the banking and broader financial sector.  To be sure, losses that arise from this source of risk---called, interest rate risk---were much less prominent in the most recent crisis relative to bank losses that stem from loan defaults---that is, credit risk---but there are ample examples of other earlier banking crises in which adverse developments in NIMs have played a significant role.  The most familiar of these from the U.S. perspective is the Savings and Loans (S\&L) crisis that started in the early 1980s with short-term interest rates rising above long-term interest rates---in part, as a result of the Volcker disinflation.  This resulted in interest expenses rising above interest income, NII and NIMs turning negative (in the thrift sector), and sizable losses and ultimately a substantial number of bank failures.  Likewise, the Nordic banking crises of the late 1980s and early 1990s (for Finland, Sweden, and Norway)---which represent three of Reinhart's and Rogoff's~(2009) ``big five'' banking crises---were associated with elevated short-term interest rates that depressed NII and NIMs and lead to sizable bank losses and failures. In the Nordic crises, high short-term interest rates resulted from these countries attempting to maintain currency pegs with the Deutschemark in the face of the tight German monetary policy that accompanied its loose fiscal policy associated with re-unification spending. Finally, there is also the U.K.'s secondary banking crisis of the early 1970s, which was associated with a sudden increase in short-term interest rates that abruptly drove up banks' costs of funds so too leading to bank losses and failures.  To be sure, large increases in short-term rates and declines in net interest income were not the \textit{only} reason for any of these crises---indeed, there was at least one major class of asset prices that collapsed in each of these crises---but short-term interest rates---and, in particular, their elevated levels relative to long-term interest rates---were significant and in most cases equivalent in importance to the other causes of the crisis.

@@

Our final motivation for examining NIMs is that despite the importance of NIMs for the financial condition of banks \textbf{[and the attention that it receives from bank equity analysts]}, the modeling of NIMs has received much less attention in the risk-modeling literature relative to the modeling of bank variables related to credit risk.  To be sure, the recent increase in the importance of stress testing has led to the development of a number of models that link NIMs to macroeconomic variables---such as, short-term Treasury rates and the slope of the Treasury yield curve---with Kovner \textit{et al}.~(2011) and Covas \textit{et al}.~(2012) being two notable examples for the U.S.  However, this research places relatively less emphasis on the conditional forecasting properties of these models, which is important given that these models are ultimately to be used to project bank losses, income, and capital given macroeconomic scenarios.\renewcommand{\baselinestretch}{1}\footnote{Covas \textit{et al}. do present some forecast evaluation results in order to demonstrate the benefits of focusing on quantile projections. However, they only report results for net chargeoffs and pre-provisioning net revenue and their focus is on density forecast evaluation.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5}

In summing up our motivation on the importance of modeling NIMs and studying interest rate risk, we would also note that the Federal Reserve chose to feature interest rate risk in one of the scenarios specified for the 2013 Dodd-Frank Act (DFA) stress tests.  In particular, for the 2013 DFA stress tests, the Federal Reserve specified the adverse scenario as one in which the Treasury yield curve shifted up and flattened.  As noted by the Federal Reserve in its (proposed) Policy Statement on scenario design, the Board varies ``the approach it uses for the adverse scenario each year so that the results of the scenario provide the most value to supervisors, in light of current condition of the economy and the financial services industry.''

In this paper we examine how well a range of time-series models can explain NIMs. We emphasize the models' pseudo-out-of-sample \textit{conditional} forecast performance. We adopt this pseudo-out-of-sample approach for the simple reason that the evaluation of models based on their in-sample forecast performance, frequently leads to the econometrician over-fitting the model within the sample so as to maximize in-sample forecast performance. We focus on conditional forecasts for the reason that scenario-based stress testing of the banking system is the underlying motivation for our analysis and it is along the dimensions of \textit{conditional} forecasts that model forecast performance is important.  Focusing on conditional forecast performance also means that we can separate the task of modeling and forecasting NIMs from that of forecasting interest rates, which is itself a sizable topic (see Duffee, 2012).

In determining which variables to include in our model of NIMs we look to the variables that have been emphasized by two existing literatures on NIMs.  The first literature emphasizes the link between interest rates of different maturities and NIMs and is more in the macro-banking tradition. The second literature focuses on the optimal margin set by banks based on considerations like the competitive structure of the lending and deposit markets in which the bank participates, the degree of managerial risk aversion, and the degree of uncertainty associated with the future path of interest rates and is much more in the micro-banking tradition.

@@

\vspace{-0.1in}
There are primarily two empirical literatures on modeling NIMs.  The first literature emphasizes the link between interest rates of different maturities and NIMs and lies more in the macro-banking tradition.  The second literature focuses on the optimal margin set by banks and lies more in the micro-banking tradition.

The papers for the U.S. by Kovner \textit{et al}. and Covas \textit{et al}. noted earlier fall into the category of papers in the macro-banking tradition, where in both cases these models link NIMs to  a measure of the slope of the yield curve and a measure of short-term market rate.  Banks' earnings from interest income reflect two key services that banks supply---maturity transformation services (that is, borrowing short and lending long) and deposit transactions services---and the returns to these services are reflected by these two interest rate variables.  The slope of the yield curve reflects the return to banks from maturity transformation and thus is expected to enter NIM models with a positive sign. The short-term market rates in NIM models reflect the fact that bank deposit rates, while typically lower (since they provide transactions services), are constrained by the zero lower bound.  As such, the short rate places an upper limit on what banks can earn from the provision of their transactions services. As a result, the short rate is expected to enter the model, also, with a positive sign.

@@

In summing up our review of other empirical analyses of NIMs a few variables stand out---in addition, to short- and long-term interest rates---as relevant for our analysis.  The degree of competition faced by the bank in the loan markets and the deposit markets in which it operates stands out as an important consideration, especially given that the sample we use in our analysis is much longer than the samples considered by Ho and Saunders, Angbazo, and Saunders and Schumacher.  Interest rate volatility, also appears important, since this has likely also changed over our sample.  The other variables that previous authors in the micro-banking tradition have added to their models in an \textit{ad hoc} way we hold-off for the time being from adding to our analysis.  This is largely because our analysis at this stage is at the aggregate level and these additional variables appear to have been added to explain variations in NIMs across firms.

@@

\vspace{-0.1in}
\noindent \textbf{1. Simple forecast combination.} We regress NIMs on each yield $r(\tau )_{t-1}$ separately: \
\[
NIM_{t}=c_{\tau }+\rho _{\tau }NIM_{t-1}+\gamma _{\tau }r(\tau )_{t-1}+{\eta
_{\tau ,t}}
\]

Each regression is use to produce an $s$ step-ahead forecast of NIMs conditional on yields on treasuries with maturity $\tau $ observed through period $t+s-1.$ This conditional forecast is denoted by $NIM_{\tau ,t+s/t}.$ The simple forecast comnination is then given by:
\[
NIM_{t+s/t}=\sum_{\tau }\frac{NIM_{\tau ,t+s/t}}{N},
\]
where $N$ is the number of maturities considered.

>>>>

\begin{document}
\renewcommand{\baselinestretch}{1}
\author{{\Large Valentin Bolotnyy} \\
Federal Reserve Board \and
{\Large Rochelle M. Edge} \\
Federal Reserve Board \and
{\Large Luca Guerrieri}\\ Federal Reserve Board} 
\title{Stressing Bank Profitability for Interest Rate Risk$^*$}

\date{\today}
\maketitle

\begin{abstract}
 Regular bank stress testing is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime.
 The form of stress testing mandated by the Dodd-Frank Act for the Federal Reserve to undertake is one based around a small number of scenarios.  This approach involves specifying macroeconomic and financial scenarios that represents stressful conditions over the time horizon of the stress test and then projecting, for each bank holding company in the stress test, its losses, income, and path of capital \textit{given that scenario}. In this paper, we focus on the path of income. In particular, we consider the ability of empirical models to translate developments in interest rates into bank net interest margins (NIMs).

@@

\noindent Key Words: Net Interest Margins, Interest Rate Risk, Forecasting, Stress Tests.

@@

\noindent JEL Classification: C53, E47, G21

\vspace{2cm}
\noindent $^*$ The views
expressed in this paper are solely the responsibility of the
authors and should not be interpreted as reflecting the views of
the Board of Governors of the Federal Reserve System or of any
other person associated with the Federal Reserve System.
Contact Information: valentin.bolotnyy@frb.gov; rochelle.m.edge@frb.gov; luca.guerrieri@frb.gov.

@@

\vspace{-0.1in}
Regular bank stress testing (and the capital planning that it implies) is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime.\renewcommand{\baselinestretch}{1}\footnote{See Governor Tarullo's November 2011 speech ``The Evolution of Capital Regulation'' for this characterization of stress testing. The two other reforms that he mentions are improvements to the traditional, individual firm approach to capital regulation and the introduction of a macroprudential component to capital regulation.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} While stress testing can take many forms---for example, being based on a small number of scenarios, being based on a large number of scenarios, or even being undertaken in reverse form to achieve a certain adverse outcome---the form of the stress tests mandated under the Dodd-Frank Act for the Federal Reserve to undertake and the form being used by the Federal Reserve for its annual Comprehensive Capital Analysis and Review (CCAR) is one based around a small number of scenarios.  This approach involves specifying a small number of macroeconomic and financial scenarios that represent stressful conditions over the time horizon of the stress test and then projecting, for each bank holding company (BHC) in the stress test, its losses, income, and path of capital \textit{given that scenario}. Accordingly, there are at least two important parts to conducting bank stress tests (i)~formulating macroeconomic and financial scenarios that should be stressful to banks and (ii)~developing econometric models that are able to translate such stressful conditions into losses, income, and capital.

We examine how well a range of time-series models can explain NIMs. We emphasize the models' pseudo-out-of-sample \textit{conditional} forecast performance. We adopt this pseudo-out-of-sample approach for the simple reason that the evaluation of models based on their in-sample forecast performance, may lead to in-sample over-fitting and poor out-of-sample performance. An explicit focus on out-of-sample performance makes forecast uncertainty explicit.
Moreover, concentrating on conditional forecasts, allows us to separate the task of modeling and forecasting NIMs from that of forecasting interest rates, which is itself a sizable topic (see Duffee, 2012).

When developing econometric models to translate stressful economic conditions into key bank financial-statement variables, it is important to consider what types of developments are likely to be stressful to banks and what particular economic variables are likely to capture these stresses. It is then critical that the models being used for stress testing be able to translate developments in these economic variables into conditional forecasts for key bank financial-statement variables.  In this paper, we examine this property for models of bank net interest margins (NIMs), defined as the ratio of net interest income (NII) to interest earning assets, where NII is the difference between a bank's interest income and interest expenses.

In determining which variables to include in our model of NIMs we look to the variables that have been emphasized by two existing strands of the literature on NIMs.  The first strand emphasizes the link between interest rates of different maturities and NIMs and is more in the macro-banking tradition. The second literature focuses on the optimal margin set by banks based on considerations like the competitive structure of the lending and deposit markets in which the bank participates, the degree of managerial risk aversion, and the degree of uncertainty associated with the future path of interest rates and is much more in the micro-banking tradition.

Our motivation for focusing on models of NIMs is fourfold. First, losses from interest rate risk have been a central feature of banking crises. Second, net interest rate margins capture income from the maturity transformation services at the center of the traditional banking model. This source of income has been lackluster given the current configuration of interest rates. Third, modelling of NIMs has received scant attention so far. Forth, the ability of producing reliable conditional forecasts for a key component of net income is a the core of the scenario-based stress testing of banks.

Bank losses that arise from banks' having to pay out more in interest on their liabilities than they are earning from their assets should not be overlooked as an important source of risk to the banking and broader financial sector.  To be sure, losses that arise from this source of risk---called, interest rate risk---were much less prominent in the most recent crisis relative to bank losses that stem from loan defaults---that is, credit risk---but there are ample examples of other earlier banking crises in which adverse developments in NIMs have played a significant role.  The most familiar of these from the U.S. perspective is the Savings and Loans (S\&L) crisis that started in the early 1980s with short-term interest rates rising above long-term interest rates---in part, as a result of the Volcker disinflation.  This resulted in interest expenses rising above interest income, NII and NIMs turning negative (in the thrift sector), and sizable losses and ultimately a substantial number of bank failures.

Likewise, the Nordic banking crises of the late 1980s and early 1990s (for Finland, Sweden, and Norway)---which represent three of Reinhart's and Rogoff's~(2009) ``big five'' banking crises---were associated with elevated short-term interest rates that depressed NII and NIMs and lead to sizable bank losses and failures. In the Nordic crises, high short-term interest rates resulted from these countries attempting to maintain currency pegs with the Deutschemark in the face of the tight German monetary policy that accompanied its loose fiscal policy associated with re-unification spending. Depressed NIMs were also at the center of the U.K.'s secondary banking crisis of the early 1970s. A sudden increase in short-term interest rates abruptly drove up banks' costs of funds,  leading to extensive bank losses and failures.  To be sure, large increases in short-term rates and declines in net interest income were not the \textit{only} reason for any of these crises---indeed, there was at least one major class of asset prices that collapsed in each of these crises---but short-term interest rates---and, in particular, their elevated levels relative to long-term interest rates---were significant and in most cases equivalent in importance to the other causes of the crisis.

@@

Despite the importance of NIMs for the financial condition of banks and despite the attention that it receives from bank equity analysts, the modeling of NIMs has received much less attention in the risk-management literature relative to the modeling of bank variables related to credit risk.  To be sure, the recent increase in the importance of stress testing has led to the development of a number of models that link NIMs to macroeconomic variables---such as, short-term Treasury rates and the slope of the Treasury yield curve---with Kovner \textit{et al}.~(2011) and Covas \textit{et al}.~(2012) being two notable examples for the U.S.  However, this research places relatively less emphasis on the conditional forecasting properties of these models, which is important given that these models are ultimately to be used to project bank losses, income, and capital given macroeconomic scenarios.\renewcommand{\baselinestretch}{1}\footnote{Covas \textit{et al}. do present some forecast evaluation results in order to demonstrate the benefits of focusing on quantile projections. However, they only report results for net chargeoffs and pre-provisioning net revenue and their focus is on density forecast evaluation.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5}

Scenario based stress tests are predicated on the ability to produce reliable conditional forecasts for key measures of profitability such as net interest margins. This kind of stress test program is not limited to the U.S. banking system. Stress testing of banks based on scenarios was undertaken during the Scandinavian banking crisis to shore up confidence in the ability of banks to withstand the crisis. Most extensively, the European Banking Authority conducted stress tests based on scenario analysis throughout the European Union in 2009, 2010, and 2011. The results of such programs can only be effective in restoring confidence from the public in the midst of a crisis to the extent that the public can be convinced of the ability of regulators to assess the consequences of their scenarios on banks. If forecast methods at the frontier of efficiency have disentangling the implications of baseline and severe stress scenarios, the ability of stress test programs to avoid panic is called in question.

%In summing up our motivation on the importance of modeling NIMs and studying interest rate risk, we would also note that the Federal Reserve chose to feature interest rate risk in one of the scenarios specified for the 2013 %Dodd-Frank Act (DFA) stress tests.  In particular, for the 2013 DFA stress tests, the Federal Reserve specified the adverse scenario as one in which the Treasury yield curve shifted up and flattened.  As noted by the Federal %Reserve in its (proposed) Policy Statement on scenario design, the Board varies ``the approach it uses for the adverse scenario each year so that the results of the scenario provide the most value to supervisors, in light of %current condition of the economy and the financial services industry.''

@@

\vspace{-0.1in}
There are two strands in the literature on modeling NIMs.  The first strand emphasizes the link between interest rates of different maturities and NIMs and lies more in the macro-banking tradition.  The second strand focuses on the optimal margin set by banks and lies more in the micro-banking tradition.

Kovner \textit{et al}. and Covas \textit{et al.} are in the macro-banking tradition. Both link NIMs to the slope of the yield curve and a short-term interest rate.  Banks' earnings from interest income reflect two key services that banks supply---maturity transformation services (that is, borrowing short and lending long) and deposit transactions services---and the returns to these services are reflected by these two interest rate variables.  The slope of the yield curve reflects the return to banks from maturity transformation and thus is expected to enter NIM models with a positive sign. The short-term market rates in NIM models reflect the fact that bank deposit rates, while typically lower (since they provide transactions services), are constrained by the zero lower bound.  As such, the short rate places an upper limit on what banks can earn from the provision of their transactions services. As a result, the short rate is expected to enter the model, also, with a positive sign.

@@

In summing up, in addition, to short- and long-term interest rates, a few variables stand out as relevant for our analysis.  The degree of competition faced by the bank in the loan markets and the deposit markets in which it operates stands out as an important consideration, especially given that the sample we use in our analysis is much longer than the samples considered by Ho and Saunders, Angbazo, and Saunders and Schumacher.  Interest rate volatility, also appears important, since this has likely also changed over our sample.  The other variables that previous authors in the micro-banking tradition have added to their models in an \textit{ad hoc} way we hold-off for the time being from adding to our analysis.  This is largely because our analysis at this stage is at the aggregate level and these additional variables appear to have been added to explain variations in NIMs across firms.

@@

We construct forecasts from a wide array of methods taken from the forecasting literature. We used data with quarterly frequency and consider a forecast horizon up to ten quarters long. We are interested in forecasts for NIMs conditional on the variables in scenario $I$, which we denote $NIM_{t+s/I}$. Here are the methods considered.

\vspace{-0.1in}
\noindent \textbf{1. Simple forecast combination.} We regress NIMs on each yield $r(\tau )_{t-1}$ separately: \
\[
NIM_{t}=c_{\tau }+\rho _{\tau }NIM_{t-1}+\sum_{j=1}^4\gamma _{\tau,j}r(\tau)_{t-j}+{\eta
_{\tau ,t}}
\]
Each regression is used to produce an $s$ step-ahead forecast of NIMs conditional on yields on treasuries r(\tau) in scenario $I$ with maturity $\tau$ observed through period $t+s-1.$ This conditional forecast is denoted by $NIM_{\tau,t+s/I}.$ The simple forecast comnination is then given by:
\[
NIM_{t+s/I}=\sum_{\tau }\frac{NIM_{\tau ,t+s/I}}{N},
\]
where $N$ is the number of maturities considered.
